|

### 2022-03-17 submission/manuscript.rmd
update after co-author comments, closes #1

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
 **Objective**: To compare different risk-based methods for optimal
 prediction of individualized treatment effects from RCTs. **Study
 Design and Setting**: We simulated RCT data using diverse assumptions
 for the average treatment effect, a baseline prognostic index of risk
 (PI), the shape of its interaction with treatment (none, linear,
 quadratic or non-monotonic) and the magnitude of treatment-related
 harms (none or constant independent of the PI). In each sample we
 predicted absolute benefit using: models with a constant relative
 treatment effect; stratification in quarters of the PI; models
 including a linear interaction of treatment with the PI; models
 including an interaction of treatment with a restricted cubic spline
 (RCS) transformation of the PI; an adaptive approach using Akaike’s
 Information Criterion. We evaluated predictive performance using root
 mean squared error and measures of discrimination and calibration for
 benefit. **Results**: The linear-interaction model and the
 RCS-interaction model outperformed the constant treatment effect
 model in many simulation scenarios. The RCS-model was optimal when
 quadratic or non-monotonic deviations from a constant treatment
 effect were stronger, and when sample size was larger. Larger sample
 size also supported the adaptive approach. Illustrations in the
 GUSTO-I trial confirmed these findings. **Conclusion**: An
 interaction between baseline risk and treatment assignment should be
 considered to improve treatment effect predictions. Non-linear 
 interactions should be considered only in larger sample sizes.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
benefit, while risk-based methods proved quite robust [@vanKlaveren2019]. Most
often, risk-modeling approaches are carried out in two steps: first a risk
prediction model is developed externally or internally on the entire RCT
population, “blinded” to treatment; then the RCT population is stratified using
this prediction model to evaluate risk-based treatment effect variation
[@Kent2010]. This two-step approach identified substantial absolute treatment
effect differences between low-risk and high-risk patients in a re-analysis of
32 large trials [@Kent2016]. However, even though estimates at the risk subgroup
level may be accurate, these estimates may need further refinement for
individual patients, especially for patients with predicted risk at the
boundaries of the risk intervals. Hence, the risk-stratified approach is useful
for exploring and presenting HTE, but is not sufficient for supporting treatment
decisions for individual patients.

To individualize treatment effects, the recent PATH statement suggested various
risk-based models including a prognostic index of baseline risk (PI) and
treatment assignment [@Kent2019; @PathEnE]. We aimed to summarize and compare
different risk-based models for predicting individualized treatment effects. We
simulated RCT settings to compare the performance of these models under
different assumptions of the relation between baseline risk and treatment. We
illustrated the different models by a case study of predicting individualized
effects of treatments for acute myocardial infarction (MI) in a large randomized
controlled trial (RCT).

@@

All risk-based methods assume that a risk prediction model is available to
assign risk predictions to individual patients. For the simulations we developed
a prediction model internally on the entire population, using a logistic
regression model with main effects for all baseline covariates and treatment
assignment. Risk predictions for individual patients were derived by setting
treatment assignment to 0. Another common approach is to derive the prediction
model solely on the control patients, however this approach has been shown to
lead to biased benefit predictions [@vanKlaveren2019; @Burke2014; @Abadie2018].

@@

For each scenario setting we performed 500 replications, within which all the
considered models were fitted. For comparing between models we simulated a
super-population of size 500,000 for each scenario. We calculated RMSE and
discrimination and calibration for benefit of the models derived in each
replication of our simulation settings within this super-population. 
## Empirical illustration

@@

The linear interaction model and the RCS-3 model both displayed very good
performance under many of the considered simulation scenarios, in contrast with
the constant relative treatment effect model. The linear interaction model was
optimal in cases with smaller sample sizes and moderately performing baseline
risk prediction models, that is, it had lower RMSE, was better calibrated for
benefit and had better discrimination for benefit, even in scenarios with strong
quadratic deviations. In scenarios with true non-monotonic deviations, the
linear interaction model was outperformed by RCS-3, especially in the presence
of true treatment-related harms. Increasing the sample size or the prediction
model’s discriminative ability favored RCS-3, especially in scenarios with
non-monotonic deviations and in the presence of treatment-related harms.

@@

The adaptive approach had adequate performance, following closely on average the
performance of the “true” model in most scenarios. With smaller sample sizes it
tended to miss the treatment-risk interactions and selected simpler models
(Supplement Section 4). This conservative behavior resulted in increased RMSE
variability in these scenarios, especially in the case of true strong linear or
non-monotonic deviations from the base case scenario. Therefore, in the case of
smaller sample sizes the simpler linear interaction model may be a safer choice
for predicting absolute benefits, if a non-constant treatment effect is
suspected.

Risk-based approaches to predictive HTE estimate treatment benefit as a function
of baseline risk. A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation
scenarios. We also considered constant moderate and strong treatment-related
harms, applied on the absolute scale to expand the range of scenarios in line
with previous work [@Glasziou1995]. In a limited set of scenarios where we
assumed the existence of true treatment-covariate interactions, our conclusions
remained unchanged. Even though the average error rates increased for all the
considered methods, due to the miss-specification of the outcome model, the
linear interaction model had the lowest error rates. RCS-3 had very comparable
performance. The constant treatment effect model often gave biased results,
espcially in the presence of moderate or strong treatment-related harms. All the
results of these simulations can be found in Supplement, Section 7. Future
simulation studies could explore the effect of more extensive deviations from
risk-based treatment effects.

In our simulations we only focused on risk-based methods, using baseline risk as
a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging from
treatment effect modeling to tree-based approaches available in more recent
literature [@Athey2019; @Lu2018;@Wager2018;@powers2018some]. Many of these
methods rely on incorporating treatment-covariate interactions in the prediction
of benefit. An important caveat of such approaches is that they may be prone to
overfitting, thus exaggerating the magnitude of the predicted benefits. In a
wide range of simulation settings, a simpler risk modeling approach was
consistently better calibrated for benefit compared to more complex treatment
effect modelling approaches [@vanKlaveren2019]. Similarly, when SYNTAX score II,
a model developed for identifying patients with complex coronary artery disease
that benefit more from percutaneous coronary intervention or from coronary
artery bypass grafting was redeveloped using fewer treatment-covariate
interactions had better external performance compared to its
predecessor[@farooq2013anatomical; @takahashi2020redevelopment].

>>>>

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
 **Objective**: To compare different risk-based methods for optimal prediction
 of individualized treatment effects. **Study Design and Setting**: We simulated
 RCT data using diverse assumptions for the average treatment effect, a baseline
 prognostic index of risk (PI), the shape of its interaction with treatment
 (none, linear, quadratic or non-monotonic), and the magnitude of
 treatment-related harms (none or constant independent of the PI). We predicted
 absolute benefit using: models with a constant relative treatment effect;
 stratification in quarters of the PI; models including a linear interaction of
 treatment with the PI; models including an interaction of treatment with a
 restricted cubic spline (RCS) transformation of the PI; an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. **Results**: The linear-interaction model and the RCS-interaction
 displayed robust performance across many simulation scenarios. The RCS-model
 was optimal when quadratic or non-monotonic deviations from a constant
 treatment effect were stronger, and when sample size was larger. The adaptive
 approach required larger sample sizes. Illustrations in the GUSTO-I trial
 confirmed these findings. **Conclusion**: An interaction between baseline risk
 and treatment assignment should be considered to improve treatment effect
 predictions.
author:
  - name: Alexandros Rekkas
    affiliation: 1
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

Risk-modeling approaches to predictive HTE analyses provide a viable option in
the absence of well-established treatment effect modifiers [@Kent2019;
@PathEnE]. In simulations, modeling of effect modifiers, i.e.
treatment-covariate interactions, often led to miscalibrated predictions of
absolute benefit, while risk-based methods proved quite robust in terms of
benefit calibration, although provided weaker discrimination of benefit in the
presence of true effect modifiers [@vanKlaveren2019]. Most often, risk-modeling
approaches are carried out in two steps: first a risk prediction model is
developed externally or internally on the entire RCT population, “blinded” to
treatment; then the RCT population is stratified using this prediction model to
evaluate risk-based treatment effect variation [@Kent2010]. This two-step
approach identified substantial absolute treatment effect differences between
low-risk and high-risk patients in a re-analysis of 32 large trials
[@Kent2016]. However, even though estimates at the risk subgroup level may be
accurate, these estimates may need further refinement for individual patients,
especially for patients with predicted risk at the boundaries of the risk
intervals. Hence, the risk-stratified approach is useful for exploring and
presenting HTE, but is not sufficient for supporting treatment decisions for
individual patients.

To individualize treatment effects, the recent PATH statement suggested various
risk-based models including a prognostic index of baseline risk (PI) and
treatment assignment [@Kent2019; @PathEnE]. In the current simulation study, we aim to summarize and compare
different risk-based models for predicting individualized treatment effects. We
simulate different relations between baseline risk and treatment effects and also consider potential harms of treatment. We
illustrate the different models by a case study of predicting individualized
effects of treatment for acute myocardial infarction (MI) in a large randomized
controlled trial (RCT).

@@

All risk-based methods assume that a risk prediction model is available to
assign risk predictions to individual patients. For the simulations we developed
a prediction model internally on the entire population, using a logistic
regression model with main effects for all baseline covariates and treatment
assignment. Baseline risk predictions for individual patients were derived by
setting treatment assignment to 0. Another common approach is to derive the
prediction model solely on the control patients, however this approach has been
shown to lead to biased benefit predictions [@vanKlaveren2019; @Burke2014;
@Abadie2018].

@@

For each scenario setting we performed 500 replications, within which all the
considered models were fitted. For comparing between models we simulated a
super-population of size 500,000 for each scenario. We calculated RMSE and
discrimination and calibration for benefit of the models derived in each
replication of our simulation settings within this super-population.

@@

The linear interaction model and the RCS-3 model both displayed very good
performance under many of the considered simulation scenarios. The linear
interaction model was optimal in cases with smaller sample sizes and moderately
performing baseline risk prediction models, that is, it had lower RMSE, was
better calibrated for benefit and had better discrimination for benefit, even in
scenarios with strong quadratic deviations. In scenarios with true non-monotonic
deviations, the linear interaction model was outperformed by RCS-3, especially
in the presence of true treatment-related harms. Increasing the sample size or
the prediction model’s discriminative ability favored RCS-3, especially in
scenarios with non-monotonic deviations and in the presence of treatment-related
harms.

@@

The adaptive approach had adequate performance, following closely on average the
performance of the “true” model in most scenarios. With smaller sample sizes it
tended to miss the treatment-risk interactions and selected simpler models
(Supplement Section 4). This conservative behavior resulted in increased RMSE
variability in these scenarios, especially in the case of true strong linear or
non-monotonic deviations from the base case scenario. Therefore, in the case of
smaller sample sizes the simpler linear interaction model may be a safer choice
for predicting absolute benefits in the presence of any suspected
treatment-related harms.

A limitation of our study is that we assumed treatment benefit
to be a function of baseline risk in the majority of the simulation
scenarios. We also considered constant moderate and strong treatment-related
harms, applied on the absolute scale to expand the range of scenarios in line
with previous work [@Glasziou1995]. In a limited set of scenarios where we
assumed the existence of true treatment-covariate interactions, our conclusions
remained unchanged. Even though the average error rates increased for all the
considered methods, due to the miss-specification of the outcome model, the
linear interaction model had the lowest error rates. RCS-3 had very comparable
performance. The constant treatment effect model often gave biased results,
espcially in the presence of moderate or strong treatment-related harms. All the
results of these simulations can be found in Supplement, Section 7. Future
simulation studies could explore the effect of more extensive deviations from
risk-based treatment effects.

In our simulations we only focused on risk-based methods, using baseline risk as
a reference in a two-stage approach to individualizing benefit
predictions. However, there is a plethora of different methods, ranging from
treatment effect modeling to tree-based approaches available in more recent
literature [@Athey2019; @Lu2018;@Wager2018;@powers2018some]. Many of these
methods rely on incorporating treatment-covariate interactions in the prediction
of benefit. An important caveat of such approaches is that they may be prone to
overfitting, thus exaggerating the magnitude of the predicted benefits. In a
wide range of simulation settings, a simpler risk modeling approach was
consistently better calibrated for benefit compared to more complex treatment
effect modelling approaches [@vanKlaveren2019]. However, whether this remains
the case in a range of empirical settings still needs to be explored Similarly,
when SYNTAX score II, a model developed for identifying patients with complex
coronary artery disease that benefit more from percutaneous coronary
intervention or from coronary artery bypass grafting was redeveloped using fewer
treatment-covariate interactions had better external performance compared to its
predecessor[@farooq2013anatomical; @takahashi2020redevelopment].
